Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
disjointness of constructors

Next, we cover the ``additive'' connectives. First, we use the
non-linear types to define \emph{indexed} versions of the additive
disjunction $\oplus$ and additive conjunction $\&$ of linear logic,
which can be thought of as linear versions of the $\Sigma$ and $\Pi$
connectives of ordinary dependent type theory, respectively. The
indexed $\&$ is defined by a $\lambda$ that brings a \emph{non-linear}
variable into scope and eliminated using projection where the index
specified is given by a non-linear term. The rules for indexed
$\oplus$ are analogous to a \emph{weak} $\Sigma$ type: it has an
injection introduction rule $\sigma$, but its elimination rule is
given by \emph{pattern matching} rather than projections
\cite{weakSigmatype}. We can define the more typical nullary and
binary versions of these connectives by using indexing over the empty
and boolean type respectively. We will freely use $0$ to refer to this
empty disjunction and $\top$ to refer to the empty conjunction, and
use infix $\oplus/\&$ for binary disjunction/conjunction.
%
In addition to the introduction and elimination rules for
$\bigamp,\bigoplus$, our development requires an additional axiom that
is not proveable from these rules: that additive conjunction
\emph{distributes} over additive disjunction, e.g., in the finitary
case that $0 \& A \cong 0$ and $(A + B) \& C \cong (A \& C) + (B \&
C)$. In its most general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse.

@@

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option'' monad
\cite{maybeMonad}. This allows for the possibility that the input
string doesn't parse, but is far too weak as a specification: we can
trivially implement a parser for any type by always returning
$\inr$. The correct notion of a parser should be one that allows for
failure, but only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$ where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to match the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

For reasons of brevity, we present only the cases related to the
linear types in \Cref{fig:semantics}. The full details are in the
appendix.\max{TODO: put the full details in the appendix}.  The
grammar for a literal $c$ has a single parse precisely when the input
string consists of the single character. The grammar for the unit
similarly has a single parse for the empty string.  A parse of the
tensor product $A \otimes B$ consists of a \emph{splitting} of the
empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
is a pair of an element of the set $X$ and a parse of $A(x)$, while
dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
that takes an $A$ parse of some other string $w'$ to a $B$ parse of
$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.

@@

\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

>>>>

Next, we cover the ``additive'' connectives. First, we use the
non-linear types to define \emph{indexed} versions of the additive
disjunction $\oplus$ and additive conjunction $\&$ of linear logic,
which can be thought of as linear versions of the $\Sigma$ and $\Pi$
connectives of ordinary dependent type theory, respectively. The
indexed $\&$ is defined by a $\lambda$ that brings a \emph{non-linear}
variable into scope and eliminated using projection where the index
specified is given by a non-linear term. The rules for indexed
$\oplus$ are analogous to a \emph{weak} $\Sigma$ type: it has an
injection introduction rule $\sigma$, but its elimination rule is
given by \emph{pattern matching} rather than projections
\cite{weakSigmatype}. We can define the more typical nullary and
binary versions of these connectives by using indexing over the empty
and boolean type respectively. We will freely use $0$ to refer to this
empty disjunction and $\top$ to refer to the empty conjunction, and
use infix $\oplus/\&$ for binary disjunction/conjunction.

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction, e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse. Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option'' monad
\cite{maybeMonad}. This allows for the possibility that the input
string doesn't parse, but is far too weak as a specification: we can
trivially implement a parser for any type by always returning
$\inr$. The correct notion of a parser should be one that allows for
failure, but only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$ where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to match the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

For reasons of brevity, we present only the cases related to the
linear types in \Cref{fig:semantics}. The full details are in the
appendix.\max{TODO: put the full details in the appendix}.  The
grammar for a literal $c$ has a single parse precisely when the input
string consists of the single character. The grammar for the unit
similarly has a single parse for the empty string.  A parse of the
tensor product $A \otimes B$ consists of a \emph{splitting} of the
empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
is a pair of an element of the set $X$ and a parse of $A(x)$, while
dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
that takes an $A$ parse of some other string $w'$ to a $B$ parse of
$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.

@@

\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}
